---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "75%"
)
```

# idpr


Overall, ‘idpr’ aims to integrate tools for the computational analysis of 
intrinsically disordered proteins within R. This package is used to identify 
known characteristics of IDPs within a sequence of interest with easily 
reported and dynamic results. Additionally, this package also includes tools
for IDP-based sequence analysis to be used in conjunction with other R 
packages. See our recently published peer-reviewed publication in 
[PLOS ONE (https://doi.org/10.1371/journal.pone.0266929)](https://doi.org/10.1371/journal.pone.0266929) 



**Please Refer to idpr-vignette.Rmd file for a detailed introduction to the**
**idpr package.**

Links to the vignettes found at the 
[Bioconductor landing page (here)](https://doi.org/doi:10.18129/B9.bioc.idpr) 
or 

## Installation

You can install the stable release version version from 
[Bioconductor](https://doi.org/doi:10.18129/B9.bioc.idpr) with:

 
```r
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("idpr")
```

Additionally, you can install the development version from 
[Bioconductor](https://bioconductor.org/packages/devel/bioc/html/idpr.html) 
with:
``` r
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# The following initializes usage of Bioc devel
BiocManager::install(version='devel')
```

Or you can install the most recent development version from 
[GitHub](https://github.com/wmm27/idpr) with:

``` r
# install.packages("devtools") #if not already installed
devtools::install_github("wmm27/idpr")
```


## Example
This is a basic example to quickly profile your protein of interest:

```{r example}
library(idpr)

P53_HUMAN <- TP53Sequences[2] #Getting a pre-loaded sequence from idpr
print(P53_HUMAN)

P53_ID <- "P04637" #Human TP53 UniProt ID

#Generates the IDP Profile:
idprofile(sequence = P53_HUMAN, 
          uniprotAccession = P53_ID, 
          proteinName = "TP53 Human", #Optional Argument
          window = 11, #Optional Argument
          pKaSet = "Lehninger", #Optional Argument
          iupredType = "redox" #Optional Argument
)
```



**Please Refer to idpr-vignette.Rmd file for a detailed introduction to the**
**idpr package.**
[Link to the Vignette (here)](https://bioconductor.org/packages/release/bioc/vignettes/idpr/inst/doc/idpr-vignette.html)


## Appendix

For use and details on 'idpr', see our peer-reviewed article published in 
[PLOS ONE (https://doi.org/10.1371/journal.pone.0266929)](https://doi.org/10.1371/journal.pone.0266929) 

### Package citation
```{r}
citation("idpr")
```

### Function citations

* Bálint Mészáros, Gábor Erdős, Zsuzsanna Dosztányi, IUPred2A: context-dependent prediction of protein disorder as a function of redox state and protein binding, Nucleic Acids Research, Volume 46, Issue W1, 2 July 2018, Pages W329–W337, https://doi.org/10.1093/nar/gky384
* Erdős, G., & Dosztányi, Z. (2020). Analyzing protein disorder with IUPred2A. Current Protocols in Bioinformatics, 70, e99. https://doi.org/10.1002/cpbi.99
* Kozlowski, L. P. (2016). IPC – Isoelectric Point Calculator. Biology Direct, 11(1), 55. https://doi.org/10.1186/s13062-016-0159-9
* Kyte, J., & Doolittle, R. F. (1982). A simple method for displaying the hydropathic character of a protein. Journal of molecular biology, 157(1), 105-132.
* Nelson, D. L., & Cox, M. M. (2017). Lehninger Principles of Biochemistry (Seventh ed.). New York, NY: W. H. Freeman and Company.
* Prilusky, J., Felder, C. E., et al. (2005). FoldIndex: a simple tool to predict whether a given protein sequence is intrinsically unfolded. Bioinformatics, 21(16), 3435-3438.
* Uversky, V. N. (2016). Paradoxes and wonders of intrinsic disorder: Complexity of simplicity. Intrinsically Disordered Proteins, 4(1), e1135015. https://doi.org/10.1080/21690707.2015.1135015
* Uversky, V. N. (2013). A decade and a half of protein intrinsic disorder: Biology still waits for physics. Protein Science, 22(6), 693-724. doi:10.1002/pro.2261
* Uversky, V. N., Gillespie, J. R., & Fink, A. L. (2000). Why are “natively unfolded” proteins unstructured under physiologic conditions?. Proteins: structure, function, and bioinformatics, 41(3), 415-427. https://doi.org/10.1002/1097-0134(20001115)41:3<415::AID-PROT130>3.0.CO;2-7


### Additional Information
```{r}
Sys.time()
Sys.Date()
R.version
```