---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "75%"
)
```

# idpr


Overall, ‘idpr’ aims to integrate tools for the computational analysis of 
intrinsically disordered proteins within R. This package is used to identify 
known characteristics of IDPs within a sequence of interest with easily 
reported and dynamic results. Additionally, this package also includes tools
for IDP-based sequence analysis to be used in conjunction with other R 
packages.

**Please Refer to idpr-vignette.Rmd file for a detailed introduction to the**
**idpr package.**
Links to the vignettes found at the 
[Bioconductor landing page](https://doi.org/doi:10.18129/B9.bioc.idpr) 


## Installation

You can install the development version from 
[Bioconductor](https://doi.org/doi:10.18129/B9.bioc.idpr) with:
``` r
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# The following initializes usage of Bioc devel
BiocManager::install(version='devel')

BiocManager::install("idpr")
```

Or you can install the development version from 
[GitHub](https://github.com/wmm27/idpr) with:

``` r
# install.packages("devtools") #if not already installed
devtools::install_github("wmm27/idpr")
```


## Example
This is a basic example to quickly profile your protein of interest:

```{r example}
library(idpr)

P53_HUMAN <- TP53Sequences[2] #Getting a preloaded sequence from idpr
print(P53_HUMAN)

P53_ID <- "P04637" #Human TP53 UniProt ID

idprofile(sequence = P53_HUMAN, #Generates the Profi
          uniprotAccession = P53_ID)
```



**Please Refer to idpr-vignette.Rmd file for a detailed introduction to the**
**idpr package.**

## Appendix

### Package citation
```{r}
citation("idpr")
```

### Additional Information
```{r}
Sys.time()
Sys.Date()
R.version
```