---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "75%"
)
```

# idpr


Overall, ‘idpr’ aims to integrate tools for the computational analysis of 
intrinsically disordered proteins within R. This package is used to identify 
known characteristics of IDPs within a sequence of interest with easily 
reported and dynamic results. Additionally, this package also includes tools
for IDP-based sequence analysis to be used in conjunction with other R 
packages.

**Please Refer to idpr-vignette.Rmd file for a detailed introduction to the**
**idpr package.**

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools") #if not already installed
devtools::install_github("wmm27/idpr")
```

## Example
This is a basic example to quickly profile your protein of interest:

```{r example}
library(idpr)

P53_HUMAN <- TP53Sequences[2] #Getting a preloaded sequence from idpr
print(P53_HUMAN)

P53_ID <- "P04637" #Human TP53 UniProt ID

idprofile(sequence = P53_HUMAN, #Generates the Profi
          uniprotAccession = P53_ID)
```


