---
title: "Charge and Hydropathy Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Charge and Hydropathy Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, 
  fig.height = 4
)
```

```{r setup}
library(idpr)
```


## Background

The composition of amino acids and the overall chemistry of Intrinsically
Disordered Proteins (IDPs) are distinctly different from that of ordered 
proteins. Each amino acid has unique chemical characteristics that can either 
favor a compact or an extended structure (Uversky, 2019) . Disorder-promoting
residues, those enriched in IDPs, are typically hydrophilic, charged, or small 
residues. Order promoting residues, those enriched in structured proteins, 
tend to be aliphatic, hydrophobic, aromatic, or form tertiary structures
(Uversky, 2013).  
Therefore, there is a distinct difference of biochemistry
between IDPs and ordered proteins. 


It was shown in Uversky, Gillespie, & Fink (2000) that both high net charge and 
low mean hydropathy are properties of IDPs. One explanation is that a high net 
charge leads to increased repulsion of residues causing an extended structure 
and low hydrophobicity will reduce the hydrophobic interactions causing reduced 
protein packing. When both average net charge and mean scaled hydropathy are 
plotted, IDPs occupy a unique area on the plot. The barrier between unfolded 
and compact proteins is:
$$<R> = 2.785 <H> - 1.151 $$ 
where 
&lt;R>
is the absolute mean 
net charge and 
&lt;H>
is the mean scaled hydropathy
(Uversky, Gillespie, & Fink, 2000).  


An alternative version of the Charge Hydropathy plot mentioned was shown
in Uversky (2016), where the average net charge is shown rather than the 
absolute value. This creates two cutoff lines. One for positively charged 
peptides: 
$$<R> = 2.785 <H> - 1.151 $$
and another for negatively charged peptides: 
$$<R> = - 2.785 <H> + 1.151 $$
(Uversky, 2016). 

This plot allows a distinction between
negative and positive proteins while preserving the information of the 
charge-hydropathy plot. 

## Installation  

The idpr package can be installed from Bioconductor with the following line of 
code. It requires the BiocManager package to be installed
```{r}
#BiocManager::install("idpr") 
```

The most recent version of the package can be installed with the following line 
of code. This requires the devtools package to be installed.

```{r}
#devtools::install_github("wmm27/idpr")
```

## Methods

Methods are originally described in Uversky, Gillespie, & Fink (2000). 
The calculations shown here are to explain how chargeHydropathyPlot() obtains
the values for graphing, and are done automatically within the function.


Mean scaled hydropathy is calculated by normalizing the Kyte and Doolittle scale
on a scale of 0 to 1, with Arg having a hydropathy of 0 and Ile having a 
hydropathy of 1, averaged by the sequence length 
(Kyte & Doolittle, 1982; Uversky, 2016). 

Net charge is calculated with the Henderson-Hasselbalch equation 
(Po & Senozan, 2001). While there is not one agreed upon set of pKa values,
the Isoelectric Point Calculator does a great job at unifying these data sets 
(Kozlowski, 2016).  The net charge is then averaged over the sequence length. 

Both values are plotted in the Charge-Hydropathy plot to identify a protein 
or proteins location in that space. 


### Example calulations

Examples will use the *H. sapiens* TP53 sequence,
acquired from UniProt (UniProt Consortium 2019)
and stored within the **idpr** package for examples.

```{r}
HUMAN_P53 <- TP53Sequences[2]
print(HUMAN_P53)
```


The first portion is calculating the mean scaled hydropathy of the sequence.
This will correspond to the value on the X-axis
```{r}
P53_msh <- meanScaledHydropathy(HUMAN_P53)
print(P53_msh)
```

The second value is the net charge. Setting averaged = TRUE will average the
calculated net charge by the sequence length.This will correspond to the 
value on the Y-axis. Total Net charge can be calculated by setting
averaged to FALSE. 

```{r}
P53_nc <- netCharge(HUMAN_P53,
                     averaged = TRUE)
print(P53_nc)
```

The chargeHydropathyFunction will automatically calculate these values and 
plot it as a point within the charge-hydropathy space according to 
Uversky, Gillespie, & Fink (2000) / Uversky (2016)

Since a ggplot is returned, the user can assign the plot to an object to 
modify aesthetics and/or add annotations. 
The example shown here will label the cooridnates on the returned plot.

```{r, include = FALSE}
P53_msh <- round(P53_msh, 3)
P53_nc <- round(P53_nc, 3)
```

```{r}
chargeHydropathyPlot(HUMAN_P53) +
  ggplot2::annotate("text",
                    x = P53_msh,
                    y = P53_nc + 0.1, #offset from point
                    label = paste("(", P53_msh, ", ", P53_nc, ")",
                                  collapse = "",
                                  sep = ""))
```




***

## Using the chargeHydropathyPlot Function

The plot above shows calculations for one protein, but the function 
can accept multiple sequences. 

The following examples will use highly similar TP53 sequences,
acquired from UniProt (UniProt Consortium 2019) that were
 stored within the **idpr** package for examples.

```{r}
TP53_Sequences <- TP53Sequences
print(TP53_Sequences)
```


```{r}
gg <- chargeHydropathyPlot(
  sequence = TP53_Sequences,
  pKaSet = "IPC_protein") 
plot(gg)
```

Since it is a ggplot, the user can add additional parameters to the plot like 
labels, custom themes, and other supported features. An example of this is
shown in "Methods".


If you do not wish to use the IPC_protein pKa set, you can change this to any 
accepted by netCharge(). 

```{r}
chargeHydropathyPlot(
  sequence = TP53_Sequences,
  pKaSet = "EMBOSS")  #Using EMBOSS pKa set
```


## Calculating Scaled Hydropathy

### Mean Scaled Hydropathy

meanScaledHydropathy is used to calculate the scaled hydropathy of a sequence.
See the introduction and methods for more details.
```{r}
 meanScaledHydropathy(HUMAN_P53)
```


### Global Hydropathy

scaledHydropathyGlobal is used to match the scaled hydropathy of 
an amino acid sequence for each residue in the sequence.

The results can be a data frame of matched values.
It will result in a data frame with 3 columns. The first column is "Position",
which indicates the numeric position of the residue in the submitted sequence. 
The second column is "AA", which indicates the amino acid residue as a single
letter. The third column is "Hydropathy", which indicates the value of scaled 
hydropathy for that residue, which was matched by the function. 
```{r}
P53_shg <- scaledHydropathyGlobal(HUMAN_P53)
head(P53_shg)
```

Or results can be returned as a plot

```{r}
scaledHydropathyGlobal(HUMAN_P53,
                       plotResults = TRUE)
```

(This is not the most aesthetically pleasing plot, so a sequenceMap from 
**idpr** is reccomended in this case for visualizations.)

```{r}
P53_shg <- scaledHydropathyGlobal(HUMAN_P53)
sequenceMap(sequence = P53_shg$AA,
            property = P53_shg$Hydropathy,
            customColors = c("chocolate1", "grey65", "skyblue3"))
```




### Local Hydrophobicity 

scaledHydropathyLocal is a function used to calculate the average 
hydrophobicity of a sequence using a sliding window. 
The results are returned as a data frame or a plot.

The scaled hydropathy is matched to each residue then the average of
each window is calculated. 

This can be helpful to identify regions of high or low hydrophobicity within
a protein. This may help with identifying IDRs as well. 


Results can be returned as a data frame of window hydropathy scores.
It will result in a data frame with 3 columns. The first column is "Position",
which indicates the numeric position of the residue in the submitted sequence. 
The second column is "Window" which indicates all the residues within the
sliding window that were used for calculations. The "CenterResidue" column 
specifies the amino acid residue as a single letter that is located in the 
center of the window and is located at the number specified by "Position"
And finally, the "WindowHydropathy" is the calculated average hydropathy for
the residues specified in the "Window" column.

```{r}
P53_shl <- scaledHydropathyLocal(HUMAN_P53,
                                 plotResults = FALSE)
head(P53_shl)
```

The results can also return a plot that shows
each window's hydropathy score along the sequence.
```{r}
scaledHydropathyLocal(HUMAN_P53,
                       plotResults = TRUE)
```


The window size can be specified as well by the "window" argument. 
**This must be an odd integer**

```{r}
scaledHydropathyLocal(HUMAN_P53,
                       window = 3,
                       plotResults = TRUE)
```





## Calculating Charge

### Net Charge
netCharge is used to calculate the net charge of a sequence.
See the introduction and methods for more details.
```{r}
netCharge(HUMAN_P53)
```

Setting averaged = TRUE will average the calculated net charge by the sequence 
length (Shown before in methods). 

```{r}
netCharge(HUMAN_P53,
          averaged = TRUE)
```


netCharge relies on the Henderson-Hasselbalch equation (via the 
hendersonHasselbalch function). Therefore the pH and pKa are critical for 
calculations.
netCharge allows for different pH values using the pH argument. 

```{r}
netCharge(HUMAN_P53,
          pH = 5.5)
netCharge(HUMAN_P53,
          pH = 7)
netCharge(HUMAN_P53,
          pH = 8)
```

There are also many pKa sets that are preloaded in **idpr**. 
pKa datasets used within this vignette are cited. See the documentation for
netCharge or pKaData within **idpr** for additional information and citations 
for available pKa sets.
Additionally, see Kozlowski (2016) for further details on pKa data sets. 

* "EMBOSS" -  (Rice, Longden, & Bleasby, 2000)
* "DTASelect"
* "Solomons"
* "Sillero"
* "Rodwell"
* "Lehninger"
* "Toseland"
* "Thurlkill"
* "Nozaki"
* "Dawson"
* "Bjellqvist"
* "ProMoST"
* "Vollhardt"
* "IPC_protein" (Kozlowski, 2016).
* "IPC_peptide" (Kozlowski, 2016).


```{r}
netCharge(HUMAN_P53,
          pKaSet = "IPC_protein")
netCharge(HUMAN_P53,
          pKaSet = "IPC_peptide")
netCharge(HUMAN_P53,
          pKaSet = "EMBOSS")

```


Alternatively, the user may supply a custom pKa dataset.
The format must be a data frame where: Column 1 must be a 
character vector of residues AND Column 2 must be a numeric vector of pKa 
values. This can be helpful if there is a data set the user prefers or if
there are noncannonical amino acids.
Here is an example using the pKa values from Wikipedia 
(Proteinogenic amino acid, n.d.).
https://en.wikipedia.org/wiki/Proteinogenic_amino_acid#Chemical_properties

```{r, include=FALSE}
charged_aa <- c("C", "D", "E", "H", "K", "R", "Y", "NH3", "COO")
aa_pKa <- c(8.55, 3.67, 4.25, 6.54, 10.40, 12.3, 9.84, 9.28, 1.99)
custom_pKa <- data.frame(AA = charged_aa,
                         pKa = aa_pKa)
```


```{r}
print(custom_pKa)

netCharge(HUMAN_P53,
          pKaSet = custom_pKa,
          includeTermini = FALSE)
```



### Global Charge Distibution

chargeCalculationGlobal is a function used to calculate the charge of
each residue, independent of other amino acids, within a sequence. 
The results are returned as a data frame (default) or a plot.

chargeCalculationGlobal accepts the same pKa and pH arguments as netCharge. 

The results can return a data frame of charge for each residue

It will result in a data frame with 3 columns. The first column is "Position",
which indicates the numeric position of the residue in the submitted sequence. 
The second column is "AA", which indicates the amino acid residue as a single
letter. The third column is "Charge", which indicates the calculated charge
for the residue at the specified pH. 
```{r}
P53_ccg <- chargeCalculationGlobal(HUMAN_P53)
head(P53_ccg)
```

The results can return a ggplot visualizing the charge distribution.
```{r}
chargeCalculationGlobal(HUMAN_P53,
                        plotResults = TRUE)
```

(This is not the most aesthetically pleasing plot, so a sequenceMap from 
**idpr** is recommended in this case for visualizations.)

```{r}
P53_ccg <- chargeCalculationGlobal(HUMAN_P53) #repeating from above
sequenceMap(sequence = P53_ccg$AA,
            property = P53_ccg$Charge,
            customColors = c("red", "blue", "grey65"))
```


The C-terminus here has a charge of ~ -2 since the function aggregates the
termini values with residue charges by default. If you wish to calculate
the termini as separate values, use sumTermini = FALSE. This will add 2 residues
to the data frame as "NH3" and "COO"

```{r}
P53_ccg <- chargeCalculationGlobal(HUMAN_P53,
                        sumTermini = FALSE)
head(P53_ccg)
```


If you wish to completely ignore the termini for calculation, set includeTermini
= FALSE. 

```{r}
P53_ccg <- chargeCalculationGlobal(HUMAN_P53,
                        includeTermini = FALSE)
head(P53_ccg)
```


### Local Charge 

chargeCalculationLocal is a function used to calculate the charge of
a sequence using a sliding window. 
The results are returned as a data frame (default) or a plot.

chargeCalculationLocal accepts the same pKa and pH arguments as netCharge. 

Unlike chargeCalculationGlobal, the chargeCalculationLocal function does not 
consider termini for calculations.

Results can be returned as a data frame of window hydropathy scores.
It will result in a data frame with 4 columns. The first column is "Position",
which indicates the numeric position of the residue in the submitted sequence. 
The "CenterResidue" column 
specifies the amino acid residue as a single letter that is located in the 
center of the window and is located at the number specified by "Position".
The "Window" column contains all the residues within the
sliding window that were used for calculating the charge.
And finally, the "windowCharge" is the calculated average charge for
the residues specified in the "Window" column.

```{r}
P53_cgl <- chargeCalculationLocal(HUMAN_P53)
head(P53_cgl)
```

Alternatively, results can be returned as a plot of each window's charge.
```{r}
chargeCalculationLocal(HUMAN_P53,
                       plotResults = TRUE)
```


The window size can be specified as well via the "window" argument. 
**This must be an odd integer**


```{r}
chargeCalculationLocal(HUMAN_P53,
                       window = 11,
                       plotResults = TRUE)
```





## References

### Packages Used

```{r}
citation("ggplot2")
```


### Citations

### References

Consortium, T. U. (2018). UniProt: a worldwide hub of protein knowledge. 
Nucleic acids research, 47(D1), D506-D515. doi:10.1093/nar/gky1049


Kozlowski, L. P. (2016). IPC – Isoelectric Point Calculator.
Biology Direct, 11(1), 55. doi:10.1186/s13062-016-0159-9

Kyte, J., & Doolittle, R. F. (1982). A simple method for
displaying the hydropathic character of a protein. Journal of molecular
biology, 157(1), 105-132. 


Po, H. N., & Senozan, N. (2001). The Henderson-Hasselbalch equation:
its history and limitations. Journal of Chemical Education, 78(11), 1499. 

Proteinogenic amino acid. (n.d.). In Wikipedia. Retrieved July 12th, 2020. 
https://en.wikipedia.org/wiki/Proteinogenic_amino_acid#Chemical_properties

Rice, P., Longden, I., & Bleasby, A. (2000). 
EMBOSS: The European Molecular Biology Open Software Suite.
Trends in Genetics, 16(6), 276-277. doi:10.1016/S0168-9525(00)02024-2


Uversky, V. N. (2013). A decade and a half of protein intrinsic disorder:
Biology still waits for physics. Protein Science, 22(6), 693-724.
doi:10.1002/pro.2261


Uversky, V. N. (2016). Paradoxes and wonders of intrinsic disorder:
Complexity of simplicity. Intrinsically Disordered Proteins, 4(1), e1135015. 
doi:10.1080/21690707.2015.1135015


Uversky, V. N. (2019). Intrinsically Disordered Proteins and Their “Mysterious” 
(Meta)Physics. Frontiers in Physics, 7(10). doi:10.3389/fphy.2019.00010


Uversky, V. N., Gillespie, J. R., & Fink, A. L. (2000). 
Why are “natively unfolded” proteins unstructured under physiologic conditions?.
Proteins: structure, function, and bioinformatics, 41(3), 415-427.
https://doi.org/10.1002/1097-0134(20001115)41:3<415::AID-PROT130>3.0.CO;2-7

### Additional Information
R Version
```{r}
R.version.string
```

System Information
```{r}
as.data.frame(Sys.info())
```

```{r}
sessionInfo()
```

```{r, results="asis"}
citation()
```










