---
title: "Sequence Map Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sequence Map Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, 
  fig.height = 4
)
```


## Introduction

One way to visualize results both within **idpr** and with data from other 
sources is the sequenceMap() function. The purpose of this function is to show 
the entire sequence and color residues based on properties. This may help 
identify important residues within a protein.

It has been identified that as few as a single amino acid is sufficient for an 
IDP to dock to its binding partner (Warren & Shechter, 2017). Therefore, it is
important to look at sequence-wide data along with the context of the primary 
sequence.  

To make a sequence map, two vectors of data are needed (typically 
in the form of a data frame). 
One vector is the amino acid sequence and another is a vector
of data to color the residues based on some property. 

Examples will use the *H. sapiens* TP53 sequence, acquired from UniProt 
(UniProt Consortium 2019)
and stored within the **idpr** package for examples.


## Installation  

The package can be installed from Bioconductor with the following line of code.
It requires the BiocManager package to be installed.
```{r}
#BiocManager::install("idpr") 
```

The most recent version of the package can be installed with the following line 
of code. This requires the devtools package to be installed.

```{r}
#devtools::install_github("wmm27/idpr")
```


## Basics of sequenceMap

```{r setup}
library(idpr) #load the idpr package
```

```{r}
P53_HUMAN <- TP53Sequences[2]
print(P53_HUMAN)
```

The values can be discrete, like the output of structuralTendency(), or 
continuous, like the output of chargeCalculationGlobal()

```{r}
tendencyDF <- structuralTendency(sequence = P53_HUMAN) 
head(tendencyDF)


chargeDF <- chargeCalculationGlobal(sequence = P53_HUMAN,
                                    includeTermini = FALSE)
head(chargeDF)

```

The output of structuralTendency() returns both the amino acid sequence in the
column 'AA' and the matched values in 'Tendency'.

The output of chargeCalculationLocal(plotResults = FALSE) 
returns both the amino acid sequence in the column 'AA' and the calculated
values in 'Charge'.
```{r}
sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency) 


sequenceMap(
  sequence = as.character(chargeDF$AA),
  property = chargeDF$Charge, #character vector
  customColors = c("blue", "red", "grey30")) 

```


## Customizations

There are multiple customization options to allow for improved graphing. 
One is the organization of the labels. 
You are able to represent the sequence with both amino acid residues and their 
location in the sequence, but you can choose one or the other (or nethier).
This is specified by the 'labelType' argument

```{r}
sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  labelType = "AA") #Only AA residue Labels

sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  labelType = "number") #Only residue numner labels
```

There is also the option of where to plot the labels. either "on" or 
"below" the sequence graphic. This is done with the "labelLocation" argument.
```{r}
sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  labelType = "number",
  labelLocation = "on") #Residue numbers printed on the sequence graphic


sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  labelType = "number",
  labelLocation = "below") #Residue numbers printed below the sequence graphic
```

The text can also be rotated, via the 'rotationAngle' argument, for ease of 
reading. This is espeically helpful for larger sequences with dense graphics.
```{r}
sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  labelType = "number",
  labelLocation = "on",
  rotationAngle = 90)
```

Also to avoid an overwhelming visual, you can specify the pattern of labels
with everyN. This number selects every Nth residue to be printed. everyN = 1 
every residue is printed. everyN = 10 every 10th residue is printed.
```{r}
sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  labelType = "number",
  labelLocation = "on",
  everyN = 1) #Every residue

sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  labelType = "number",
  labelLocation = "on",
  everyN = 2) #Every 2nd (or every other)

sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  labelType = "number",
  labelLocation = "on",
  everyN = 10) #Every 10th residue is printed
```

You can also change the number of residues on each line with nbResidues. 
This helps improve visualization based on the sequence size. Visualizing a 
larger sequence will likely benefit from a larger nbResidues value.
```{r}
sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  labelType = "number",
  labelLocation = "on",
  nbResidues = 15) #15 residues each row

sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  labelType = "number",
  labelLocation = "on",
  nbResidues = 45,
  rotationAngle = 90) #45 residues each row
```


You can also specify colors for discrete values using a vector of colors. This
is done with the "customColors" argument.
```{r}
sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  customColors = c("#999999", "#E69F00", "#56B4E9")) 
```

Continuous variables custom colors are specified with a vector in the order of 
"High value", "Low  Value", "Middle Value". Here the order is high = purple, 
low = pink, and middle = light grey
```{r}
sequenceMap(
  sequence = as.character(chargeDF$AA),
  property = chargeDF$Charge, 
  customColors = c("purple", "pink", "grey90")
  ) 
```

Since the output is a ggplot, the visualization is able to be assigned to an
object and additional features can be added and annotated. The example below 
will annotate a metal binding residue and the region that the P53 protien binds 
DNA. These annotations and locations were retrieved from UniProt 
(UniProt Consortium 2019).
```{r}
library(ggplot2)

ggSequence <- sequenceMap(
  sequence = tendencyDF$AA,
  property = tendencyDF$Tendency,
  nbResidues = 40,
  customColors = c("#999999", "#E69F00", "#56B4E9"))

    # Adding Annotations of DNA Binding from UniProt
ggSequence <- ggSequence +
              annotate("segment",
                       x = 21,
                       xend  = 40.5,
                        y = 8.05,
                        yend = 8.05,
                       color = "#FF3562",
                       size = 1.5) + 
              annotate("segment",
                       x = 1,
                       xend  = 12.5,
                        y = 3.05,
                        yend = 3.05,
                       color = "#FF3562",
                       size = 1.5) + 
              annotate("segment",
                       x = 1,
                       xend  = 40.5,
                        y = c(7:4) + 0.05,
                        yend = c(7:4) + 0.05,
                       color = "#FF3562",
                       size = 1.5) +
               annotate("segment",
                        x = 34,
                        xend = 36,
                        y = 0.65,
                        yend = 0.65,
                        color = "#FF3562",
                       size = 1.5) +
              annotate("text",
                       x = 36.35, 
                       y = 0.65,
                       label = "= DNA Binding",
                       size = 3.5,
                       hjust = 0)
  
  
# Adding a plot title
ggSequence <- ggSequence +
              labs(title = "P53 Structural Tendency") +
              theme(plot.title = element_text(hjust = 0.5,
                                              vjust = 2.5))
# Adding point and text annotations
ggSequence <- ggSequence +
              geom_point(aes(x = 2.5, #column 2
                             y = 4), #row 4
                         shape = 8,
                         show.legend = FALSE,
                         inherit.aes = FALSE) +
              annotate("text",
                       x = 4.5, 
                       y = 4.3,
                       label = "Metal Binding",
                        size = 3)

plot(ggSequence)
```




## Getting Coordinates

Since the sequenceMap() function plots the sequence as a grid, adding additional
features to the correct location may be difficult without knowing the location
of the residues on the map.

To solve this, sequenceMapCoordinates() will return the row (y value) and the 
column (x value) as a data frame for each residue visualized with sequenceMap(). 
The puporse of this is to make adding annotations easier and customizable.


As shown before, nbResidues deterines how many residues will be on each
row. Make sure nbResidues is equal to the value used in sequenceMap(). 

To get the coordinates, the amino acid sequence must be supplied. The output
is a data frame with each reside and its column and row values. These can 
then be used to make annotations on the map.


```{r}
coord_DF <- sequenceMapCoordinates(P53_HUMAN,
                                   nbResidues = 40)
head(coord_DF)
```


## Sequence Plot

The funtions for calculating charge and scaled hydropathy and the iupred
functions all have plotting options. The plotting for these are done with the
sequencePlot() function to have a uniform aesthetic. If you wish to make a plot
with custom values, sequencePlot() can still be used.

Horizontal lines are provided by hline (the dashed line in the middle) and
propertyLimits (the solid lines at the top and bottom).

Colors can be dynamic by setting dynamicColor equal to the value of the color.

Since the result is ggplot element, a user can add custom annotations, 
change the axes, or modify the theme with any acceptable ggplot functions.

```{r}
exampleDF <- chargeCalculationGlobal(P53_HUMAN,
                                     includeTermini = FALSE)

#Making a sequence plot
sequencePlot(
  position = exampleDF$Position,
  property = exampleDF$Charge)

#Adding a dynamic colors based on the property values and horizontal lines
sequencePlot(
position = exampleDF$Position,
  property = exampleDF$Charge,
hline = 0.0,
propertyLimits = c(-1.0, 1.0),
dynamicColor = exampleDF$Charge,
customColors = c("red", "blue", "grey50"),
customTitle = "Charge of Each Residue / Terminus")
```



## References

### Packages Used

```{r}
citation("ggplot2")
```


### Citations


Kozlowski, L. P. (2016). IPC – Isoelectric Point Calculator. Biology Direct, 
11(1), 55. doi:10.1186/s13062-016-0159-9


Kulkarni, P., & Uversky, V. N. (2018). Intrinsically Disordered Proteins: 
The Dark Horse of the Dark Proteome. PROTEOMICS, 18(21-22), 1800061. 
doi:10.1002/pmic.201800061


UniProt Consortium. (2019). UniProt: a worldwide hub of protein knowledge. 
Nucleic acids research, 47(D1), D506-D515. 

Uversky, V. N. (2019). Intrinsically Disordered Proteins 
and Their 
“Mysterious” (Meta)Physics. Frontiers in Physics, 7(10).
doi:10.3389/fphy.2019.00010


Warren, C., & Shechter, D. (2017). Fly Fishing for Histones:
Catch and Release by Histone Chaperone Intrinsically Disordered Regions
and Acidic Stretches. J Mol Biol, 429(16), 2401-2426. 
doi:10.1016/j.jmb.2017.06.005





### Additional Information
R Version
```{r}
R.version.string
```

System Infomation
```{r}
as.data.frame(Sys.info())
```

```{r, results="asis"}
citation()
```



